Outline flowCore paper

### 2008-06-25 paper/manuscript.tex
updates
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32424 bc3139a8-67e5-0310-9ffc-ced21a209358

The computational tools we have developed are distributed in the R 
software language (www.r-project.org) as the Bioconductor
(www.bionductor.org) package flowCore. flowCore is a freely available, highly
functional and extensible FCM data analysis platform that 
enables researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. Among the issues
addressed by flowCore are computationally efficient data structures and
a range of specialized methods for compensation, transformation, and gating. 
flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.  In this paper, we describe the functionality and methods underlying flowCore, 
and provide examples of their use.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform has proven to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. 
%% NLM: that's probably the important point: platform for innovation rather than for standardization
Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

Traditionally, the majority of FCM experiments have been analyzed
manually by direct data inspection in one or two dimensions, or by
very basic comparisons of summary statistics. We believe that these
approaches do not fully address the highly complex nature of FCM data,
in particular, they discard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. In addition, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection has become time consuming and labor intensive, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that close collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. 
%% I find the previous sentence redundant - 
%% They may be applied on raw or transformed parameters
%% EAS: I don't believe the previous sentence is required.
This enables the encoding of a simple analytical workflow, in a way
that allows us to reconstruct the analysis programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing workflows and to communicate with
every other FCM tool that adheres to the proposed
standard. Adherence to standards also plays a critical role in the ability
of new methods based on flowCore to find there way back into the
standard practices for FCM data analysis.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches. By taking the
burden of data management from the programmer, and by providing well-defined 
APIs and a structured class
hierarchy, it is possible to readily test new ideas and to easily
extend the framework's functionality. %Both the underlying R and
%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
%development projects, thus community involvement in the developing
%process is highly appreciated.

\section*{Basic data structures}
\subsection*{Flow Frame}
flowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other FCM data analysis tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flow frame. All relevant information for the collection
are stored along with the matrix of raw data values and can be
accessed programmatically. Most commonly, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flow frame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process that all share a common Application Program Interface (API).

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
that depend on automated sample acquisition using a plate-based platform. 
%% EAS: I don't think the previous sentence is required. Does it 
%% matter if the flowSet is ordered or not?

%However, the flow set structure can be
%readily extended to also accommodate such spatial information
%% EAS: The spatial aspect of plateCore is really secondary to having 
%% an easy, consistent way to manage and access the mulitplex annotation.
%% Also, it's important to have a standard format for the annotation if
%% we're trying to encourage collaboration. I haven't done it yet, but I 
%% intend to look into
%% into stealing the annotation implementation from CellHTS2 and
%% using it in plateCore.
However, the flow set structure can be readily extended to incorporate the
potentially complex annotation associated with large FC-HCS experiments. Access
to flowCore data structures and methods has already enabled the development of
new Bioconductor pacakges, such as plateCore, that faciliate the analysis of
plate-based FC-HCS data and make handling the annotation and processing
the data easier.

%\textit{(plateCore reference, Perry, Errol?)}.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{The flowCore frame work. For each
    experiment, FCS files, phenotypic and meta data is stored in a
    flow set. Each flow frame of a flow set corresponds to one FCS
    file. All basic operations (e.g., compensation, transformation,
    gating) can be applied on both single flow frames or entire flow
    set.}}
\end{figure}

@@

%\begin{figure}
%\centering
%\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
%\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
%    packages: dedicated packages have been developed that utilize
%    flowCore’s data structures. flowViz is a data visualization
%    package; flowQ provides data quality assessment methods; flowUtils
%    contains interfaces for flow data, including xml parsers:
%    flowClust provides automated gating by model-based clustering.}}
%\end{figure}

>>>>

The computational tools we have developed are distributed in the R
software language (www.r-project.org) as the Bioconductor
(www.bionductor.org) package flowCore. flowCore is a freely available,
highly functional and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent analysis. Among the
issues addressed by flowCore are computationally efficient data
structures and a range of specialized methods for compensation,
transformation, and gating.  flowCore runs on Windows, Mac OS X and
Linux/Unix operating systems. Conceptually, the underlying data
structures that build the core of our software can be implemented in
any programming language, and we chose the R language for its obvious
advantages in statistical programming. In this paper, we describe the
functionality and methods underlying flowCore, and provide examples of
their use.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project aims to
provide R software modules for biolocial and clinical data analysis.
It is widely used to process data in various applications, spanning a
diversity of research fields from genomics to proteomics to cell
biology \citep{gentleman2006bos}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data obtained. We rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second lesson being the importance of a
modular architecture that allows for many developers to extend and use
the underlying infrastructure and to combine in complex workflows.

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
However, recent developments in high-throughput FCM are shifting the
focus of interest away from single-tube based measurements towards
large and complex experimental designs with dozens of covariates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Many of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products makes it impractical to integrate them into
analysis pipelines.

Traditionally, the majority of FCM experiments have been analyzed by
manual data inspection in one or two dimensions, or by very basic
comparisons of summary statistics. We believe that these approaches,
in addition to being extremely expensive and labor intensive, do not
fully address the highly complex nature of FCM data; in particular,
they discard many of the fundamental aspects of the data, such as its
underlying distribution and its high-dimensional nature. Furthermore,
the subjective character of manual analyses are a major obstacle to
reproducibility. For FC-HCS data, unassisted manual inspection is too
time consuming, and robust statistical methods need to be developed to
point investigators to interesting aspects of the data, or to
potential problems. While the expert knowledge of immunologists and
researchers remains crucial for the understanding of FCM data, we
believe that collaboration with other research fields such as
statistics and computer science can greatly improve the relevance of
FCM in today's high-throughput paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate
Recommendation (CR) represents a proposal on how to form unambiguous
XML-based gate definitions, which can facilitate the interchange and
validation of data between different software packages. Gates may be
ordered into a hierarchical structure to describe a gating
strategy. This enables the encoding of a simple analytical workflow,
in a way that allows us to reconstruct the analysis programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing workflows and to communicate with every
other FCM tool that adheres to the proposed standard. Adherence to
standards also plays a critical role in the ability of new methods
based on flowCore to find there way back into the standard practices
for FCM data analysis. Note that the data strucures defined in
flowCore aim to model large sets of FCM data in a natural and coherent
way; they are not specific to the R language and can be implemented in
any other programming language. Ideally, different software programs
would communicate on the level of these data structures rather than a
high-level markup language.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches. By
taking the burden of data management from the programmer, and by
providing well-defined APIs and a structured class hierarchy, it is
possible to readily test new ideas and to easily extend the
framework's functionality.

\section*{Basic data structures}
\subsection*{flowFrame}
flowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very close to
that adopted by other Bioconductor packages. All information from a
single FCS file, i.e., the collection of events and the accompanying
metadata, is stored in one single container. We call the structure
that hold this data a flowFrame, and raw data values as well as
associated metadata can be accessed programmatically. Most commonly,
these are descriptors of the stains used in the experiment and the
respective measurement channels, information about compensation
performed at the instrument side and any additional keywords the user
deems to be important to annotate the data. A number of quality checks
are performed during the creation of a flowFrame to ensure data
integrity.

\subsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. One can envision a flowSet to be a container for multiple
flowFrames along with relevant information associated with each
individual frame such as descriptions of the cell sample or the
treatment to which the sample was subjected. They manage the
consistent application of operations on the individual flow frames and
shift the burden of keeping score of the annotation from the user to
the infrastructure, thus reducing the risk of common errors
(\textit{e.g.}, mixups of sample labels). The flowSet structure can be
readily extended to incorporate the potentially complex annotation
associated with large FC-HCS experiments. Access to flowCore data
structures and methods has already enabled the development of new
Bioconductor packages, such as plateCore, that faciliate the analysis
of plate-based FC-HCS data and make handling the annotation and
processing the data easier.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
    individual flowFrames), phenotypic and metadata are stored in a
    flowSet. Each flowFrame in a flowSet corresponds to one FCS
    file. All basic operations (e.g., compensation, transformation,
    gating) can be applied to either single flowFrames or a flowSet
    simulataneously.}}
\end{figure}

@@

%\begin{figure}
%\centering
%\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
%\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
%    packages: dedicated packages have been developed that utilize
%    flowCoreœôòùs data structures. flowViz is a data visualization
%    package; flowQ provides data quality assessment methods; flowUtils
%    contains interfaces for flow data, including xml parsers:
%    flowClust provides automated gating by model-based clustering.}}
%\end{figure}
